---
title: "R final(GILOMA)"
author: "Pradip Basnet"
date: "`r Sys.Date()`"
output: pdf_document
---

## loading the necessary packages

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(readr)
library(plotly)
library(GGally)
library(vcd)
library(beeswarm)
```


```{r}
data <- read_csv("TCGA_InfoWithGrade.csv")
head(data)
#view(data)
```


```{r}
str(data)
glimpse(data)
```


```{r}
#view(data)
summary(data)
```

checkig whether data has missing values or not

```{r}
sum(is.na(data))
#view(data)
```

```{r}
colSums(is.na(data))
```
```{r}
# Binning the age column into categories
data$Age_group <- cut(data$Age_at_diagnosis, breaks = c(0, 30, 50, 70, 100), labels = c("0-30", "30-50", "50-70", "70-100"))
#view(data)
```



"Grade"            "Gender"           "Age_at_diagnosis" "Race"             "IDH1"             "TP53"             "ATRX"            
 [8] "PTEN"             "EGFR"             "CIC"              "MUC16"            "PIK3CA"           "NF1"              "PIK3R1"          
[15] "FUBP1"            "RB1"              "NOTCH1"           "BCOR"             "CSMD3"            "SMARCA4"          "GRIN2A"          
[22] "IDH2"             "FAT4"             "PDGFRA"           "Age_group"        "mutation_count"  


```{r}
# Add a feature representing the total number of mutated genes
data$mutation_count <- rowSums(data[, c("IDH1","TP53","ATRX","PTEN","EGFR","CIC","MUC16","PIK3CA","NF1","PIK3R1","FUBP1","RB1","NOTCH1","BCOR","CSMD3","SMARCA4","GRIN2A","IDH2","FAT4","PDGFRA")])
data
```


```{r}
# Convert Grade column: 0 to "LGG" and 1 to "GBM"
data$Grade <- ifelse(data$Grade == 0, "LGG", "GBM")

# Verify the changes
table(data$Grade)  # Check the distribution of LGG and GBM in the Grade column
#view(data)
```

## convert the gender columns into male and female 

```{r}
# Check unique values in a specific column
unique_values <- unique(data$Race)
print(unique_values)

table(data$Race)
```
```{r}
# Modify the Race column based on the provided categories
data$Race <- case_when(
  data$Race == 0 ~ "White",
  data$Race == 1 ~ "Asian",
  data$Race == 2 ~ "Black / African American",
  data$Race == 3 ~ "Not Reported",
  TRUE ~ "Other"  # Add this to handle any unexpected values
)

# View the counts of each race
table(data$Race)

```


```{r}
data$Gender <- ifelse(data$Gender == 0, "Male", "Female")
table(data$Gender)
```


```{r}
# Check the number of missing values in the 'Age_at_diagnosis' column
missing_values <- sum(is.na(data$Race))
print(missing_values)

```


```{r}
# Round the Age_at_diagnosis column to the nearest whole number
data$Age_at_diagnosis <- round(data$Age_at_diagnosis)

# Verify the changes
head(data$Age_at_diagnosis)

```



```{r}
colnames(data)
#view(data)
```


```{r}
# Assuming your cleaned data is in a data frame called 'cleaned_data'
write.csv(data, 'cleaned_data.csv', row.names = FALSE)

```


## Descriptive Analysis

```{r}
# Summarize dataset to understand basic characteristics
summary(data)
#view(data)
```


```{r}
# Select mutation columns
mutation_columns <- c('IDH1', 'TP53', 'ATRX', 'PTEN', 'EGFR', 'CIC', 'MUC16', 
                      'PIK3CA', 'NF1', 'PIK3R1', 'FUBP1', 'RB1', 'NOTCH1', 
                      'BCOR', 'CSMD3', 'SMARCA4', 'GRIN2A', 'IDH2', 'FAT4', 
                      'PDGFRA')

# Summing mutations across all individuals for each gene
mutation_frequencies <- data %>% 
  select(all_of(mutation_columns)) %>%
  summarise_all(sum) %>%
  gather(Gene, Frequency)

# Plot bar plot
p <- ggplot(mutation_frequencies, aes(x = reorder(Gene, -Frequency), y = Frequency)) +
  geom_bar(stat="identity", fill="cornflowerblue") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title = "Mutation Frequency by Gene", x = "Gene", y = "Number of Mutations")
ggplotly(p)
```






```{r}
# Calculate mutation count per individual
data$mutation_count <- rowSums(data[mutation_columns])

# Plot boxplot by Age Group
ggplot(data, aes(x = Age_group, y = mutation_count)) +
  geom_boxplot(fill="lightgreen") +
  labs(title = "Mutation Count Distribution by Age Group", x = "Age Group", y = "Number of Mutations")

```







```{r}
# Violin plot comparing mutation count distribution across Age Group
p <- ggplot(data, aes(x = Age_group, y = mutation_count, fill = Age_group)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Mutation Count Distribution by Age Group", x = "Age Group", y = "Mutation Count") +
  theme_minimal()
ggplotly(p)
```



```{r}
# Density plot of mutation count across the entire dataset
p <- ggplot(data, aes(x = mutation_count)) +
  geom_density(fill="blue", alpha=0.5) +
  labs(title = "Density Plot of Mutation Count", x = "Mutation Count", y = "Density") +
  theme_minimal()
ggplotly(p)
```




```{r}
# Boxplot with jitter to show mutation count distribution across tumor grade
p <- ggplot(data, aes(x = as.factor(Grade), y = mutation_count, color = Grade)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.5) +
  labs(title = "Mutation Count by Tumor Grade", x = "Tumor Grade", y = "Mutation Count") +
  theme_minimal()
ggplotly(p)
```



```{r}
# Improve plot appearance
p <- ggplot(data, aes(x = mutation_count)) +
  geom_histogram(binwidth = 1, fill="#3498db", color="#ecf0f1") + # Updated color scheme
  facet_grid(Age_group ~ Gender) + # Facets by Age Group and Gender
  labs(title = "Mutation Count Distribution by Age Group and Gender", 
       x = "Mutation Count", 
       y = "Frequency") +
  theme_minimal() +
  theme(
    text = element_text(family = "Arial", color = "#2c3e50"), # Set text family and color
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5), # Centered bold title
    axis.title = element_text(size = 14, face = "bold"), # Bold axis titles
    strip.text = element_text(size = 12, face = "bold"), # Bold facet labels
    panel.grid.major = element_blank(), # Remove major grid lines
    panel.grid.minor = element_blank(), # Remove minor grid lines
    panel.background = element_rect(fill = "white", colour = "grey"), # White background with grey border
    legend.position = "none" # Hide legend if not necessary
  )

# Convert to an interactive plot using ggplotly
ggplotly(p)

```





```{r}
# Create a mosaic plot for joint distribution of Tumor Grade and IDH1 Mutation
mosaic(~Grade + IDH1, data = data, 
       shade = TRUE, 
       legend = TRUE, 
       labeling_args = list(set_varnames = c(Grade = "Tumor Grade", IDH1 = "IDH1 Mutation")),
       main = "Mosaic Plot: Tumor Grade vs IDH1 Mutation")

```



```{r}
# Swarm plot for mutation count across tumor grade
beeswarm(mutation_count ~ as.factor(Grade), data = data, 
         pch = 16, col = rainbow(6), main = "Swarm Plot: Mutation Count by Tumor Grade", 
         xlab = "Tumor Grade", ylab = "Mutation Count")
```




```{r}
library(ggplot2)
library(dplyr)

# Simulated example data
set.seed(123)
data <- data.frame(
  Tumor_Grade = factor(sample(c("LGG", "GBM"), 100, replace = TRUE)),
  Mutation_Type = factor(sample(c("IDH1", "TP53", "ATRX"), 100, replace = TRUE)),
  Mutation_Count = rnorm(100, mean = 10, sd = 3)
)

# Summarizing data to get mean mutation count for each combination
interaction_data <- data %>%
  group_by(Tumor_Grade, Mutation_Type) %>%
  summarise(Mean_Mutation_Count = mean(Mutation_Count), .groups = 'drop')

# Creating the interaction plot
ggplot(interaction_data, aes(x = Tumor_Grade, y = Mean_Mutation_Count, color = Mutation_Type, group = Mutation_Type)) +
  geom_line() +
  geom_point() +
  labs(title = "Interaction Plot", x = "Tumor Grade", y = "Mean Mutation Count", color = "Mutation Type")

```




```{r}
# Group data by Grade and Mutation status
mutation_by_grade <- data %>%
  group_by(Grade) %>%
  summarise_at(mutation_cols, ~ sum(!is.na(.)))

# Convert data to long format for plotting
mutation_by_grade_long <- mutation_by_grade %>%
  gather(Mutation, Count, -Grade)

# Stacked bar chart of mutations by tumor grade
ggplot(mutation_by_grade_long, aes(fill=Mutation, y=Count, x=Grade)) +
  geom_bar(position="stack", stat="identity") +
  labs(title="Mutation Distribution by Tumor Grade", x="Tumor Grade", y="Count") +
  theme_minimal()

```


```{r}
mutation_cols <- names(data)[5:ncol(data)]

# Count mutations by gender
mutation_by_gender <- data %>%
  group_by(Gender) %>%
  summarise(across(all_of(mutation_cols), ~ sum(. == 1), .names = "count_{.col}"))  # Assuming mutations are encoded as 1

# Convert to long format
mutation_by_gender_long <- pivot_longer(mutation_by_gender,
                                        cols = starts_with("count"),
                                        names_to = "Mutation",
                                        names_prefix = "count_",
                                        values_to = "Count")

# Plot
ggplot(mutation_by_gender_long, aes(fill = Gender, y = Count, x = Mutation)) +
  geom_bar(position = "dodge", stat = "identity") +
  labs(title = "Mutation Prevalence by Gender", x = "Mutation", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```









## Distribution of Clinical Variables

```{r}
# 2.1: Grade Distribution (Bar chart)
p <- ggplot(data, aes(x = factor(Grade), fill = Grade)) +
  geom_bar() +
  labs(title = "Distribution of Grade", x = "Grade", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("cornflowerblue", "orange"))  +
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(p)

```


## # 2.2: Age at Diagnosis Distribution (Histogram)
```{r}
p <- ggplot(data, aes(x = Age_at_diagnosis)) +
  geom_histogram(binwidth = 5, fill = "cornflowerblue", color = "white") +
  labs(title = "Distribution of Age at Diagnosis", x = "Age at Diagnosis", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplotly(p)
```


```{r}
p <- ggplot(data, aes(x = Age_group)) +
  geom_bar(binwidth = 5, fill = "cornflowerblue", color = "white") +
  labs(title = "Distribution of Age at Diagnosis", x = "Age Groups", y = "Count") +
  theme_minimal()
ggplotly(p)
```

## Gender Distribution (Pie chart) 

```{r}
# Calculate gender distribution with counts and percentages
gender_dist <- data %>%
  group_by(Gender) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot the pie chart
ggplot(gender_dist, aes(x = "", y = Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Distribution of Gender") +
  theme_minimal() +
  geom_text(aes(label = paste0(Count, " (", round(Percentage, 1), "%)")),
            position = position_stack(vjust = 0.5)) +  # Place text in the middle of each pie slice
  theme(axis.text.x = element_blank(),  # Remove x-axis labels
        axis.ticks = element_blank(),   # Remove saxis ticks
        plot.title = element_text(hjust = 0.5))  # Center the title

```

```{r}
# First calculate the counts and percentages
race_dist <- data %>%
  group_by(Race) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Plot the bar chart with different colors and percentage labels on top of each bar
p <- ggplot(race_dist, aes(x = factor(Race), y = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            vjust = -0.5, size = 4) +  # Display percentage on top of the bars
  labs(title = "Distribution of Race", x = "Race", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

# Convert ggplot to plotly for interactivity
ggplotly(p)

```


## Mutational Landscape

```{r}
mutation_cols <- c("IDH1", "TP53", "ATRX", "PTEN", "EGFR", "CIC", "FUBP1", "RB1", "NOTCH1", "BCOR", "CSMD3", "SMARCA4", "GRIN2A", "IDH2", "FAT4", "PDGFRA")
mutation_freq <- colSums(data[ , mutation_cols] == 1)

mutation_df <- data.frame(Gene = names(mutation_freq), Frequency = mutation_freq)

p <- ggplot(mutation_df, aes(x = reorder(Gene, -Frequency), y = Frequency)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  labs(title = "Frequency of Genetic Mutations", x = "Gene", y = "Frequency") +
  theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5))
  coord_flip()
ggplotly(p)
```

## Co-mutation Analysis (IDH1 + TP53) ----> not sure grna ki nai vnera ailea lai....!!!

## Feature Relationships

```{r}
#Relationship Between Age and Grade (Box plot)
p <- ggplot(data, aes(x = factor(Grade), y = Age_at_diagnosis, fill = factor(Grade))) +
  geom_boxplot() +
  labs(title = "Age Distribution by Grade", x = "Grade", y = "Age at Diagnosis") +
  theme_minimal()
ggplotly(p)
```

## Relationship Between Gender and Grade (Stacked bar chart)

```{r}
p <- ggplot(data, aes(x = factor(Grade), fill = factor(Gender))) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of Gender Across Grades", x = "Grade", y = "Proportion") +
  theme_minimal()
ggplotly(p)
```





## Relationship Between Mutation Count and Grade (Box plot)

```{r}
ggplot(data, aes(x = factor(Grade), y = mutation_count, fill = Grade)) +
  geom_boxplot() +
  labs(title = "Mutation Count by Grade", x = "Grade", y = "Mutation Count") +
  theme_minimal()
```


```{r}
p <- ggplot(data, aes(x = factor(Race), y = mutation_count, fill = Race)) +
  geom_boxplot() +
  labs(title = "Mutation Count by Race", x = "Race", y = "Mutation Count") +
  theme_minimal()
ggplotly(p)

```


## Heatmap of Mutations Across Patients

```{r}
library(GGally)

# Pair plot with key variables
ggpairs(data, columns = c("Age_at_diagnosis", "mutation_count", "Grade"))

```


```{r}

ggplot(data, aes(x = factor(Grade), y = Age_at_diagnosis, fill = Grade)) +
  geom_violin() +
  labs(title = "Violin Plot of Age vs Grade", x = "Grade", y = "Age at Diagnosis") +
  theme_minimal()

```

```{r}
# Create a new data frame for mutational burden
mutation_burden <- data %>%
  group_by(Grade) %>%
  summarise(mutation_count = sum(mutation_count))

# Stacked bar chart for mutational burden by Grade
ggplot(mutation_burden, aes(x = factor(Grade), y = mutation_count, fill = factor(Grade))) +
  geom_bar(stat = "identity") +
  labs(title = "Mutational Burden by Grade", x = "Grade", y = "Total Mutation Count") +
  theme_minimal()

```

```{r}
# Create a grouped bar plot for gender and age groups
ggplot(data, aes(x = Age_group, fill = Gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Comparison of Age Groups Across Gender", x = "Age Group", y = "Count") +
  theme_minimal()

```



```{r}
library(ggridges)

ggplot(data, aes(x = mutation_count, y = Grade, fill = Grade)) +
  geom_density_ridges() +
  labs(title = "Ridge Plot of Mutation Count by Grade", x = "Mutation Count", y = "Grade") +
  theme_ridges()

```


```{r}
ggplot(data, aes(x = Age_at_diagnosis)) +
  geom_density(fill = "skyblue", alpha = 0.7) +
  labs(title = "Density Plot of Age at Diagnosis", x = "Age at Diagnosis", y = "Density") +
  theme_minimal()

```



```{r}
library(networkD3)

nodes <- data.frame("name" = unique(c(data$Grade, data$Race)))
links <- data.frame(source = match(data$Grade, nodes$name) - 1,
                    target = match(data$Race, nodes$name) - 1,
                    value = rep(1, nrow(data)))

sankeyNetwork(Links = links, Nodes = nodes, Source = "source", Target = "target", Value = "value", NodeID = "name")

```



```{r}
library(pheatmap)

mutations <- data[, c("IDH1", "TP53", "ATRX", "PTEN", "EGFR", "CIC", "FUBP1", "RB1", "NOTCH1", "BCOR", "CSMD3", "SMARCA4", "GRIN2A", "IDH2", "FAT4", "PDGFRA")]
pheatmap(cor(mutations), cluster_rows = TRUE, cluster_cols = TRUE)

```





```{r}
library(GGally)
ggparcoord(data, columns = 5:ncol(data), groupColumn = "Grade") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Parallel Coordinate Plot for Mutations and Grade")

```


